---
title: "Transform Data 2"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

```{r packages}
library("tidyverse")
library("babynames")
```


## Activity 1

Use summarise() to compute three statistics about the data:

- The smallest (minimum) year in the dataset
- The largest (maximum) year in the dataset
- The total number of children represented in the data

```{r}

```


## Activity 2

Extract the rows where name is "Khaleesi". Then use summarise() and a summary functions to find:

- The total number of children named Khaleesi
- The first year Khaleesi appeared in the data

```{r}

```


## Practice Data

```{r}
pollution <- tibble(city = c("New York", "New York", "London", "London", "Beijing", "Beijing"),
                    size = c("large", "small", "large", "small", "large", "small"),
                    amount = c(23, 14, 22, 16, 121, 121))
```


## Activity 3

Use group_by() and summarise() to calculate the number of male and female babies born in each year. 

```{r}

```


## Activity 4

On the storms data set, calculate the maximum wind speed and minimum pressure for each hurricane in each year, and arrange the summary in descending order of wind speed.

```{r}

```


## Activity 5

Building on the previous code, calculate the average hurricane wind_max for each year. Which year had the most intense hurricanes, on average?

```{r}

```


## Activity 6

What is the average diameter of the area experiencing hurricane strength winds (hu_diameter) for each category of hurricane (category)?

```{r}

```


## Activity 7a

- For each vore, calculate the average of only the columns that start with "sleep" in the msleep data set.
- Hint: use na.rm = TRUE

```{r}

```

## Activity 7b

- For each vore, calculate the average of only the numeric columns in the msleep data set.
- Hint: use na.rm = TRUE

```{r}

```


## Activity 7c

- For each vore, calculate the average of all columns in the msleep data set.
- Hint: use na.rm = TRUE

```{r}

```


## **Not discussed in class:** how to apply several _different_ functions (with possibly different arguments) to multiple columns.

Notes:

- The different functions should be provided in a list.
- Names that you provide for the summary functions become suffixes on the old column names (in the example below, sleep_total_avg, sleep_total_min, etc.).
- The `.x` argument is a placeholder for the underlying data to summarize.

```{r}
msleep %>% 
  group_by(vore) %>% 
  summarise(across(starts_with("sleep"), 
                   list(avg = ~ mean(.x, na.rm = TRUE),
                        min = ~ min(.x, na.rm = TRUE),
                        max = ~ max(.x, na.rm = TRUE))))
```

## Activity 8

In babynames, try to recreate the prop column using a grouped mutate (call it "new_prop"). Specifically, divide each row's n by the total number of n for that sex and year.

Why are the values slightly different?


```{r}

```


## Activity 9A

Using msleep, determine how much each species' log body weight differs from the average log body weight for its order.


```{r}

```


## Activity 9B

Using baby names, add a rank column to each name for each year and sex. What were the top 10 ranked boys names in 2015, and what were their ranks?


```{r}

```


## Activity 10a

When building statistical models, it's often useful to standardize numeric variables by converting to z-scores:
z = (observed value - mean)  / standard deviation.

Using msleep, use mutate() to convert log(bodywt) to a z-score for each row in the data set.

Plot the data using a violin plot and stat_summary()

```{r}

```


## Activity 10b

Now do the same, but standardize within groups by using group_by(vore).

```{r}

```