---
title: "Tidy Data"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

## Setup

```{r packages}
library("tidyverse")
library("glue")
library("ggrepel")
library("rcartocolor")
library("knitr")

theme_set(theme_minimal())
```

You will need the development version of the `tidyr` package to use the new reshaping functions.

_Run only once:_

```
devtools::install_github("tidyverse/tidyr")
```

***


## Make some practice data

```{r}
cases <- tibble(country = c("FR", "DE", "US"),
                `2011` = c(7000, 5800, 15000),
                `2012` = c(6900, 6000, 14000),
                `2013` = c(7000, 6200, 13000))

kable(cases)
```

```{r}
pollution <- tibble(city = c("New York", "New York", "London", "London", "Beijing", "Beijing"),
                    size = c("large", "small", "large", "small", "large", "small"),
                    amount = c(23, 14, 22, 16, 121, 121))

kable(pollution)
```


## Activity 1

Plan (e.g., draw on paper) how the `cases` data set would look if it had the same values grouped into three columns: **country**, **year**, **n**.

```
Nothing to put here.
```


## Activity 2

Use `pivot_long()` to reorganize `table4a` into three columns: **country**, **year**, and **cases**.

```{r}
# Your R code goes here
```


## Activity 3

Plan (e.g., draw on paper) how the `pollution` data set would look if it had the same values grouped into three columns: **city**, **large**, **small**.

```
Nothing to put here.
```

## Activity 4

Use `pivot_wide()` to reorganize `table2` into four columns: **country**, **year**, **cases**, and **population**.

```{r}
# Your R code goes here
```

## Activity 5

Remember the pay gaps data set?

```{r}
# You will probably have to modify this...
earnings <- read_csv(here::here("06_viz-relationships-change", "data", 
                                "school-earnings.csv"))

kable(head(earnings))
```

Let's visualizing pay disparities between men and women that graduated from 21 elite U.S. universities using a paired-value bubble plot. Specifically, we want to plot: 

- Women's salaries on the x-axis
- Men's salaries on the y-axis
- The size and color of the point show the size of the pay gap
- A 1-to-1 line showing gender parity, to provide context

Reshape the `earnings` data so that it has the following columns (variables): `School`, `Gap`, `Women` and `Men`. Then, fill in the blanks to set aesthetic mappings to produce the plot.

**Remove `eval = FALSE` from the two chunks below**

```{r, eval = FALSE}
earnings_wide <- earnings %>% 
  __________ # Your pivot statement
```

#### Make the plot

```{r, fig.width = 9, fig.height = 7, eval = FALSE}

# Pay gap breaks for legend
gap_breaks <- seq(0, 60, by = 10)
gap_labs <- glue("${gap_breaks}k")

# Pay amount breaks for breaks for axis
pay_breaks <- seq(0, 150, by = 50)
pay_labs <- glue("${pay_breaks}k")

# Paired scatterplot over 1-to-1 diagonal
ggplot(earnings_wide, aes(______________)) +
  geom_abline(aes(_________), size = 1.5, color = "gray85") +
  geom_point(aes(_______________), alpha = 0.8) +
  geom_text_repel(aes(__________),
                  point.padding = 0.5, show.legend = FALSE) +
  coord_equal() +
  expand_limits(x = 0, y = 0) +
  scale_x_continuous(breaks = pay_breaks, labels = pay_labs) +
  scale_y_continuous(breaks = pay_breaks, labels = pay_labs) +
  scale_size_area(name = "Pay Gap", breaks = gap_breaks, 
                  labels = gap_labs, max_size = 8) +
  scale_color_carto_c(name = "Pay Gap", palette = "RedOr",
                      breaks = gap_breaks, labels = gap_labs,
                      guide = "legend") +
  labs(x = "Women's Mid-career Salary",  y = "Men's Mid-career Salary")
```
