---
title: "Spreadsheets"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      fig.align = "center", 
                      fig.height = 6, 
                      fig.width = 8)
```


```{r}
library("tidyverse")
library("knitr")
library("readxl")
library("here")
library("janitor")
```


# Sticker Wealth

**Set your path here:**

```{r}
stickers_path <- here("09_data-import/spreadsheets/PosidFazioCordes_StickerRich_Excel-Dataverse.xlsx")
```

We're going to read in some messy data from an Excel file that accompanies a journal article about how children's altruistic behavior is shaped by their "wealth" in terms of the number of stickers they have: [Being Sticker Rich: Numerical Context Influences Children’s Sharing Behavior](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0138928). The data are [archived in a data repository](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/LHAQ5W), but they can also be downloaded from the course website.

As a first attempt, let's just read in the file using `read_excel()` with no other options provided.

```{r}
stickers <- read_excel(stickers_path)
```

Let's look at that mess.

```{r}
stickers
```

That second row of data is wonky, and we can just get rid of it. There's a nice `dplyr` function called `slice()` for choosing rows based on their position in the table. Here's a demonstration:

```{r, eval = FALSE}
# Get row 2 only
slice(stickers, 2)

# Get rows 1 to 3
slice(stickers, 1:3)

# Get the last row only with the function n(), which is the number of rows in the table
slice(stickers, n())

# Get rows 2 to last row
slice(stickers, 2:n())

# Another way of writing the same thing is with negative indices (for rows to to remove)
slice(stickers, -1)
```


Another issue is that some of the column names have completely ridiculous names like `absolute number of stickers given (Conditions 1 or 3: Out of 12; Conditions 2 or 4: Out of 30)`.

```{r}
names(stickers)
```

If we enjoyed wasting time, we could fix that by manually renaming each column. Instead, let's use the R package `janitor`, which provides a handy function `clean_names()` for making nice names using a variety of rules. By default, it makes names using [snake_case](https://en.wikipedia.org/wiki/Snake_case), but you can find all the other options in the [documentation](https://rdrr.io/cran/snakecase/man/to_any_case.html), including such delightful naming rules as "big_camel" (NamesLikeThis) and "screaming_snake" (NAMES_LIKE_THIS).

```{r}
clean_stickers <- clean_names(stickers)
names(clean_stickers)
```

## Finished Product: Stickers

Read in the stickers data in one fell swoop using piped functions. Your chain should be 5 lines long:

1. Start with path to the data
2. Call `read_excel()`
3. Slice out the bad row
4. Rename the long column
5. Clean the names using `clean_names()`


```{r, eval = FALSE}
# Remove eval = FALSE from this chunk

stickers <- # Your R code goes here
```


<br>

***

<br>

# Messy Artifacts

In this activity we're going to import data from a messy spreadsheet of mass measurements on archeological artifacts including lithics, ceramics, and molluscs. I found the data in a public respository for an [Archeological Data Analysis Workshop](https://github.com/benmarwick/SAA_R_workshop_2017) by MAtthew Harris and Ben Marwick.

Go ahead and open the file `messy_ktc_data.xlsx`. Look upon the single worksheet with tables nested within tables and despair. This file violates practically every guideline in Broman and Woo's ["Data organization in spreadsheets"](https://kbroman.org/dataorg/).

We have three goals here:

1. Read in the subtable labeled "Ceramic data" and put it in an object called `ceramic`.
2. Read in the subtable labeled "lithic data" and put it in an object called `lithic`.
3. Read in all the mollusc data organized into 17 (?!?!!) subtables and put it all together in an object called `mollusc`.

**Set your path here:**

```{r}
artifacts_path <- here("09_data-import/spreadsheets/messy_ktc_data.xlsx")
```


The first two are easy. Simply provide the cell range to the `read_excel()` function.

```{r eval = FALSE}
# Remove eval = FALSE from this chunk

ceramic <- artifacts_path %>% 
  # read_excel() code

glimpse(ceramic)
```


```{r eval = FALSE}
# Remove eval = FALSE from this chunk

lithic <- artifacts_path %>% 
  # read_excel() code

glimpse(lithic)
```

Mollusc is a bit more complicated, but still not so bad. If we simply provide the cell range:

```{r}
mollusc <- artifacts_path %>% 
  read_excel(range = "L18:X217")

glimpse(mollusc)
```

Everything is treated as a text column because the column headings for each subtable are intermixed with the data values. To fix this, write a sequence of commands in which you:

1. Start with path to the data.
2. Call read_excel() on the appropriate range and use the `col_types` argument to specify the column type for each column.
3. Filter out all the rows that correspond with column headings.
4. Clean the names using `clean_names()` from the `janitor` package

```{r eval = FALSE}
# Remove eval = FALSE from this chunk

mollusc <- # Your R code goes here

glimpse(mollusc)
```

Finally, let's make a histogram to visualize the mass distributions of bivalves and gastropods from our sample.

```{r eval = FALSE}
# Remove eval = FALSE from this chunk

ggplot(filter(mollusc, element %in% c("Bivalve", "Gastropod")), 
       aes(x = mass_g + 0.1)) +            # Add a little bit to mass to avoid log(0) problem
  geom_histogram(color = "white", 
                 fill = "thistle") + 
  geom_hline(yintercept = 0) +             # Make nice black baseline at zero
  scale_x_log10(labels = scales::comma) +  # Log-transform x-axis, and use comma lables
  facet_wrap(~element, ncol = 1) +         # Arrange facets in single column
  coord_cartesian(expand = FALSE) +        # Don't add extra space around extremes of the plot
  theme_minimal(14) +
  labs(x = "Mass (g)", y = "Count", 
       title = "Mass of molluscs", 
       caption = "Data from https://github.com/benmarwick/aswr/tree/master/inst/data")
```

