---
title: "Transform Data"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

```{r packages}
library("tidyverse")
library("babynames")
```

## babynames

```{r}
babynames
```


## Activity 1

Alter the code to select just the `n` column:

```{r}
select(babynames, name, prop)
```

## Quiz

Which of these is NOT a way to select the `name` and `n` columns together?

```{r}
select(babynames, -c(year, sex, prop))
select(babynames, name:n)
select(babynames, starts_with("n"))
select(babynames, ends_with("n"))
```

## Activity 2

See if you can use the logical operators to manipulate babynames to show:

- All of the names where prop is greater than or equal to 0.08
- All of the children named "Sea"
- All of the names that have a missing value for n (Hint: this should return an empty data set).

```{r}
filter(babynames, is.na(n))
```

## Activity 3

Use logical operators to alter the code below to return only the rows that contain:

* Girls named Sea  
* Names that were used by exactly 5 or 6 children in 1880  
* Names that are one of Acura, Lexus, or Yugo

```{r}
filter(babynames, name == "Sea" | name == "Anemone")
```

## Activity 4

Arrange babynames by n, and add name as a second (tie breaking) variable by which to arrange. Can you tell what the smallest value of `n` is?

```{r}

```

## Activity 5

Use `desc()` to find the names with the highest prop.
Then, use `desc()` to find the names with the highest n.

```{r}

```

## Activity 6

Use `%>%` to write a sequence of functions that: 

1. Filter babynames to just the girls that were born in 2015  
2. Select the `name` and `n` columns  
3. Arrange the results so that the most popular names are near the top.

```{r}

```

## Activity 7

- Use `mutate()` and `min_rank()` to rank each row in babynames from largest n to smallest n, and filter to extract the top 10 ranking names in the entire data set.
- Do the same but use prop rather than n to rank the names.
- What differences do you see in these rankings?

```{r}

```

## Activity 8

Use `mutate()` and `lag()` to find the year-to-year change with the largest increase in `n` for your name and sex.

```{r}

```
