---
title: "Two-table Operations"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

```{r packages}
library("tidyverse")
library("nycflights13")
```


## Practice Data

```{r}
band <- tibble(name = c("Mick", "John", "Paul"),
               band = c("Stones", "Beatles", "Beatles"))

instrument <- tibble(name = c("John", "Paul", "Keith"),
                     plays = c("guitar", "bass", "guitar"))
```


## Activity 1

Which airlines had the largest arrival delays? Complete the code below. 


```{r eval = FALSE}
flights %>%
  drop_na(arr_delay) %>%
  __________________ %>%
  group_by(_________) %>%
  __________________ %>%
  arrange(__________)
```


## Activity 2

- Use flights and airports to compute the average arr_delay by destination airport (names only, not codes), and the number of flights that this is based on.

- Arrange by average delay, from worst to best.

```{r}
# Your R code goes here
```


## Activity 3

- Can you figure out how to get the full name of both the origin airport (origin) and the destination airport (dest) in the flights table?

- Do as before, but group by both origin name and destination name.

```{r}
# Your R code goes here
```


## Activity 4

How many airports in airports are serviced by flights originating in New York (i.e. flights in our dataset?) 

```{r}
# Your R code goes here
```


## Activity 5

What are the unique dest codes present in the flights table that have no corresponding information in the airports table?.

```{r}
# Your R code goes here
```